SummaryStats

### 2023-03-26 analysis/code/Exploration.Rmd
#3 ggsave most exploration plots

```{r observations across time}
ggplot(data) + 
  geom_bar(aes(x=year))
```

@@

```{r number of lineages across time}
yearly_plot <- terrs %>% 
  ggplot(aes(x=year))+
  geom_bar()

# count distinct terr_ids that show up in a century
centuries_plot <- terrs %>% 
  mutate(century = year - year %% 100) %>% # round year down to century
  distinct(century, terr_id) %>% # else counts territory-year observations
  ggplot(aes(x=century))+
  geom_bar()

plot_grid(yearly_plot, centuries_plot, labels = c("Yearly", "Centuries"))
```

@@

# create plot with outliers
size_time <- terrs_centuries %>% 
  ggplot() + 
  geom_boxplot(aes(x = century, y = count_cities))

# create plot with no outliers
size_time_no_outliers <- terrs_centuries %>% 
  ggplot() + 
  geom_boxplot(aes(x = century, y = count_cities), outlier.shape = NA) +
  coord_cartesian(ylim = quantile(terrs_centuries$count_cities, c(0, 0.93)))

# plot side by side
plot_grid(size_time, size_time_no_outliers, labels = c("Full plot", "Zoom"))
```

@@

CityExtPlot <- function(full_data, bins){
  plt <- full_data %>% 
    filter(extinction == 1) %>% 
    ggplot(aes(x=year))+
    geom_histogram(binwidth = bins)
  return(plt)
}

plot_grid(CityExtPlot(data, 1), CityExtPlot(data, 10),
          CityExtPlot(data, 30), CityExtPlot(data, 50))

@@

TerrExtPlot <- function(terrs, bins){
  plt <- terrs %>% 
    filter(extinction == 1) %>% 
    ggplot(aes(x=year))+
    geom_histogram(binwidth = bins)
  return(plt)
}

plot_grid(TerrExtPlot(terrs, 1), TerrExtPlot(terrs, 10),
          TerrExtPlot(terrs, 30), TerrExtPlot(terrs, 50))

@@

size_ext <- terrs_life %>% 
  ggplot() + 
  geom_boxplot(aes(x = extinction_ever, y=max_size))

size_ext_no_outliers <- terrs_life %>% 
  ggplot() + 
  geom_boxplot(aes(x = extinction_ever, y=max_size), outlier.shape = NA) +
  coord_cartesian(ylim = quantile(terrs_life$max_size, c(0, 0.93)))

plot_grid(size_ext, size_ext_no_outliers, labels = c("Full plot", "Zoom"))
```

@@

```{r lifespan of lineages}
terrs_life %>% 
  ggplot()+
  geom_histogram(aes(x=lifespan))
```

@@

```{r}
#mod_lifespan_size <- lm(lifespan ~ log(max_size), data=terrs_life)
#summary(mod_lifespan_size)

@@

Left: The lifespan of a lineage is weakly correlated to its maximum lifetime size.
Right: This honestly doesn't say much. What it tries to say is that lineages may be less likely to go extinct when they are larger.

@@

plot_integ <- function(data, width, startyear=0, endyear=2000){
  cities <- data %>% 
    filter(year >= startyear & year <= endyear) %>% 
    mutate(period = year - year %% width) %>% 
    mutate(period = factor(period)) %>%
    group_by(period, city_id) %>% 
    summarise(mean_integ = mean(count_cities))

  plot <- cities %>% 
    ggplot() +
    geom_boxplot(aes(x = period, y = mean_integ))

@@

plot_grid(integ_50y, integ_1600_25y, integ_1800_10y,
          align="v", ncol=1, labels = c("Full", "1600+", "1800+"))
```

@@

>>>>

```{r observations across time}
ggplot(data) + 
  geom_bar(aes(x=year)) +
  labs(
    title="Number of observations across time"
  )

ggsave("obs_yearly.png", device="png", path="analysis/output/exploration")
```

@@

```{r number of lineages across time}
yearly_plot <- terrs %>% 
  ggplot(aes(x=year))+
  geom_bar()+
  labs(
    title="Distinct lineages across time"
  )

# count distinct terr_ids that show up in a century
centuries_plot <- terrs %>% 
  mutate(century = year - year %% 100) %>% # round year down to century
  distinct(century, terr_id) %>% # else counts territory-year observations
  ggplot(aes(x=century))+
  geom_bar()+
  labs(
    title="Distinct lineages in each century"
  )

plot_grid(yearly_plot, centuries_plot)
ggsave("terrs_yearly_cents.png", device="png", path="analysis/output/exploration")
```

@@

# create plot with outliers
size_time <- terrs_centuries %>% 
  ggplot() + 
  geom_boxplot(aes(x = century, y = count_cities))+
  labs(
    title="Territory size across time",
    y = "Mean city count"
  )

# create plot with no outliers
size_time_no_outliers <- terrs_centuries %>% 
  ggplot() + 
  geom_boxplot(aes(x = century, y = count_cities), outlier.shape = NA) +
  coord_cartesian(ylim = quantile(terrs_centuries$count_cities, c(0, 0.93)))+
    labs(
    title="Zoom",
    y = "Mean city count"
  )

# plot side by side
plot_grid(size_time, size_time_no_outliers)
ggsave("terrsize_cents_boxplot.png", device="png", path="analysis/output/exploration")

@@

CityExtPlot <- function(full_data, bins, title){
  plt <- full_data %>% 
    filter(extinction == 1) %>% 
    ggplot(aes(x=year))+
    geom_histogram(binwidth = bins)+
    labs(
      title=title
    )
  return(plt)
}

plot_grid(CityExtPlot(data, 1, "Cities affected by extinction, yearly"), CityExtPlot(data, 10, "10-year bins"),
          CityExtPlot(data, 30, "30-year bins"), CityExtPlot(data, 50, "50-year bins"))

ggsave("cities_ext.png", device="png", path="analysis/output/exploration")
```

@@

TerrExtPlot <- function(terrs, bins, title){
  plt <- terrs %>% 
    filter(extinction == 1) %>% 
    ggplot(aes(x=year))+
    geom_histogram(binwidth = bins)+
    labs(
      title=title
    )
  return(plt)
}

plot_grid(TerrExtPlot(terrs, 1, "Lineage extinctions, yearly"), TerrExtPlot(terrs, 10, "10-year bins"),
          TerrExtPlot(terrs, 30, "30-year bins"), TerrExtPlot(terrs, 50, "50-year bins"))
ggsave("terrs_ext.png", device="png", path="analysis/output/exploration")

@@

size_ext <- terrs_life %>% 
  ggplot() + 
  geom_boxplot(aes(x = extinction_ever, y=max_size))+
  coord_flip()+
  labs(
    title="Maximum size of lineages with recorded extinction"
  )

size_ext_no_outliers <- terrs_life %>% 
  ggplot() + 
  geom_boxplot(aes(x = extinction_ever, y=max_size), outlier.shape = NA) +
  coord_flip(ylim = quantile(terrs_life$max_size, c(0, 0.915)))+
  labs(
    title="Zoom"
  )

plot_grid(size_ext, size_ext_no_outliers, ncol = 1)
ggsave("ext_max_size.png", device="png", path="analysis/output/exploration")

@@

```{r lifespan of lineages}
terrs_life %>% 
  ggplot()+
  geom_histogram(aes(x=lifespan))+
  labs(
    title="Distribution of lineage lifespan",
    subtitle="(Last year minus first year that lineage shows up in the city data)"
  )

ggsave("lifespan_histo.png", device="png", path="analysis/output/exploration")
```

@@

```{r eval=FALSE, include=FALSE}
#mod_lifespan_size <- lm(lifespan ~ log(max_size), data=terrs_life)
#summary(mod_lifespan_size)

@@

(The below refers to the output of a deactivated cell.)
Left: The lifespan of a lineage is weakly correlated to its maximum lifetime size.
Right: This honestly doesn't say much. What it tries to say is that lineages may be less likely to go extinct when they are larger.

@@

plot_integ <- function(data, width, startyear=1300, endyear=1918){
  cities <- data %>% 
    filter(year >= startyear & year <= endyear) %>% 
    mutate(period = year - year %% width) %>% 
    mutate(period = factor(period)) %>%
    group_by(period, city_id) %>% 
    summarise(mean_integ = mean(count_cities))

  plot <- cities %>% 
    ggplot() +
    geom_boxplot(aes(x = period, y = mean_integ))+
    labs(
      title=sprintf("Integration of cities (period mean), %i-%i", startyear, endyear)
    )

@@

plot_grid(integ_50y, integ_1600_25y, integ_1800_10y,
          align="v", ncol=1)
ggsave("city_integ_boxplots.png", device="png", path="analysis/output/exploration")
```

@@

# figure out how to unify the colour legends before saving this!
